# E-Processes and Anytime-Valid Inference {#sec-eprocesses}

```{r}
#| label: setup-ch4
#| include: false
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

In @sec-safe-tests, we repeatedly used the fact that safe tests remain valid under optional stopping, so a researcher may monitor the e-value during data collection and stop once it crosses $1/\alpha$ while still controlling the type-I error at level $\alpha$. In that chapter, however, this was treated as a property of particular tests such as the safe z-test, safe t-test, safe $2 \times 2$ test, and safe logrank test. This chapter steps back from those examples and develops the general sequential machinery that makes such guarantees possible.

The two central ideas are **e-processes**, which extend e-values from a single analysis time to an evolving stream of evidence, and **confidence sequences**, which extend fixed-sample confidence intervals to procedures that remain valid under continuous monitoring. Together, they provide the general framework for anytime-valid inference.

## The sequential setting {#sec-sequential-setting}

The tests in @sec-safe-tests were applied to data that arrived sequentially. At each time step $n$, a new observation $X_n$ was observed and the evidence was updated. The key feature of optional stopping is that the decision to stop may depend on the information accumulated so far, but not on future observations. To state this precisely, we need a small amount of sequential vocabulary [@ramdas2024hypothesis, Chapter 7].

A **filtration** $(\mathcal{F}_n)_{n \geq 0}$ is an increasing sequence of sub-$\sigma$-algebras of $\mathcal{F}$:
$$
\mathcal{F}_0 \subseteq \mathcal{F}_1 \subseteq \mathcal{F}_2 \subseteq \cdots.
$$
Intuitively, $\mathcal{F}_n$ represents the information available by time $n$. In the canonical case, $\mathcal{F}_0 = \{\emptyset, \Omega\}$ and $\mathcal{F}_n = \sigma(X_1, \ldots, X_n)$, the $\sigma$-algebra generated by the first $n$ observations, so $\mathcal{F}_n$ then contains exactly the yes-or-no statements that can be answered from the data observed so far, such as whether the sample mean is positive or whether the running e-value has already crossed a rejection threshold. More generally, $\mathcal{F}_n$ may contain further information available by time $n$, such as treatment assignments, external randomisation, or observations from other data streams.

A random variable $\tau$ taking values in $\{0, 1, 2, \ldots\} \cup \{\infty\}$ is a **stopping time** with respect to $(\mathcal{F}_n)$ if the event $\{\tau \leq n\}$ belongs to $\mathcal{F}_n$ for every $n \geq 0$. In words, whether the experiment has stopped by time $n$ must be decidable using only the information available by time $n$, never future observations. The convention $\tau = \infty$ means that the analyst never stops.

**Example.** In the safe z-test of @sec-safe-z at $\alpha = 0.05$, one possible rule is to stop and reject as soon as the running e-value reaches the threshold $1/\alpha = 20$. The corresponding stopping time is
$$
\tau = \inf\{n \geq 1 : E_n \geq 20\},
\qquad
\inf\emptyset := \infty.
$$
This is a stopping time with respect to $\mathcal{F}_n = \sigma(X_1, \ldots, X_n)$, because after observing the first $n$ data points we can determine whether the threshold has been crossed at or before time $n$.

By contrast, a rule such as "stop at time $n$ if the next observation $X_{n+1}$ exceeds 2" is not a stopping time with respect to $(\mathcal{F}_n)$, because it depends on information that is not yet available at time $n$.

A process $(Y_n)_{n \geq 0}$ is **adapted** to $(\mathcal{F}_n)$ if $Y_n$ is $\mathcal{F}_n$-measurable for each $n$. Intuitively, this means that the value of the process at time $n$ can be computed from the information available by time $n$. All the running e-value paths in @sec-safe-tests are adapted because the value at time $n$ is a function of the first $n$ observations.

## Test supermartingales and e-processes {#sec-test-supermg}

In @sec-evalue-definition, we defined an e-variable as a nonnegative random variable with expectation at most 1 under the null. The sequential analogue replaces this single random variable by a process that evolves over time as new data arrive. The key requirement is that this running measure of evidence should remain properly calibrated under the null, not only at one fixed sample size but throughout the monitoring process [@ramdas2024hypothesis].

::: {#def-test-supermartingale}
## Test supermartingale [adapted from @ramdas2024hypothesis, Definition 7.3(i)]

A nonnegative process $(M_n)_{n \geq 0}$ adapted to $(\mathcal{F}_n)$ is a **test supermartingale** for $\mathcal{P}$ if, for every $P \in \mathcal{P}$,
$$
\mathbb{E}_P[M_0] \leq 1 \qquad \text{and} \qquad \mathbb{E}_P[M_{n+1} \mid \mathcal{F}_n] \leq M_n \quad \text{for all } n \geq 0.
$$
If, in addition,
$$
\mathbb{E}_P[M_{n+1} \mid \mathcal{F}_n] = M_n \quad \text{for all } n \geq 0 \text{ and all } P \in \mathcal{P},
$$
the process is a **test martingale**. Throughout this thesis, we use the common normalisation $M_0 = 1$ almost surely.
:::

Throughout this chapter, equalities and inequalities involving conditional expectations are understood to hold $P$-almost surely for each $P \in \mathcal{P}$.

The supermartingale condition says that, under the null, the process cannot be expected to increase from one step to the next, conditional on everything observed so far. Intuitively, once the first $n$ observations have been seen, the current value $M_n$ already reflects all evidence available at that time, so under the null there is no systematic way to grow this evidence further in expectation. In the betting language from @sec-betting, the process tracks Skeptic's wealth, and the condition is the same one that made Forecaster willing to offer the bet in the first place: under the null, Forecaster expects not to lose money on the next round, given the information currently available [@ramdas2024hypothesis].

A closely related notion captures the property that is most relevant for sequential testing.

::: {#def-eprocess}
## E-process [adapted from @ramdas2024hypothesis, Definition 7.3(ii)]

A nonnegative process $(E_n)_{n \geq 0}$ adapted to $(\mathcal{F}_n)$ is an **e-process** for $\mathcal{P}$ if, for every $P \in \mathcal{P}$ and every stopping time $\tau$ with respect to $(\mathcal{F}_n)$,
$$
\mathbb{E}_P[E_\tau] \leq 1,
$$
where $E_\infty = \limsup_{n \to \infty} E_n$.
:::

An e-process is therefore a process whose value at any stopping time is itself a valid e-variable. This is the formal meaning of optional-stopping safety, so even if the analyst decides to stop in a data-dependent way, as long as the rule is a valid stopping time, the resulting value can still be thresholded at $1/\alpha$ without invalidating type-I error control [@ramdas2024hypothesis].

The distinction between the two notions is useful. A test supermartingale is defined through a one-step conditional expectation inequality, whereas an e-process is defined directly through validity at every stopping time. Thus test supermartingales give a local, step-by-step calibration condition, while e-processes express the global property that sequential testing actually relies on. The following proposition makes the link precise.

::: {#prp-supermg-eprocess}
## Test supermartingales are e-processes [@ramdas2024hypothesis]

Every test supermartingale for $\mathcal{P}$ is an e-process for $\mathcal{P}$.
:::

This follows from Definition 7.3 and Fact 7.5 of @ramdas2024hypothesis, which imply that every test supermartingale satisfies the optional-stopping property at arbitrary stopping times. Hence, for every stopping time $\tau$ with respect to $(\mathcal{F}_n)$,
$$
\mathbb{E}_P[M_\tau] \leq \mathbb{E}_P[M_0] \leq 1
\qquad \text{for all } P \in \mathcal{P}.
$$
Thus the stopped value $M_\tau$ is a valid e-variable under every $P \in \mathcal{P}$, so $(M_n)$ is an e-process. In other words, it is the one-step conditional supermartingale inequality, rather than fixed-time calibration alone, that makes a test supermartingale safe under optional stopping.

The converse need not hold in full generality: there are testing problems for which useful e-processes exist even though no nontrivial test supermartingales are available, so e-processes can form a strictly broader class; see @ramdas2023gametheoretic and @ramdas2024hypothesis for discussion. For the present thesis, however, this distinction is mostly technical rather than substantive: the safe tests we use are all constructed as test supermartingales, and the e-process property then follows automatically.

## Building e-processes from conditional e-variables {#sec-conditional-evalues}

The running e-value paths in @sec-safe-tests were all constructed multiplicatively. This reflects a general principle: sequential evidence is often built from a sequence of one-step factors, where each new factor is chosen using the information available so far and then evaluated on fresh data. The right one-step notion is a **conditional e-variable**.

This definition is naturally motivated by the betting interpretation of sequential testing. At each step, before the next observation is seen, the analyst chooses a nonnegative payoff based on the past. Under the null, its conditional expected value given the past must be at most 1. In the language of @grunwald2024safe, this is a conditional e-variable given $\mathcal{F}_{i-1}$; in the language of @ramdas2024hypothesis, it is the one-step bet underlying a test supermartingale.

::: {#def-conditional-evalue}
## Conditional e-variable [adapted from @grunwald2024safe; @ramdas2024hypothesis]

A nonnegative random variable $E_i$ is a **conditional e-variable** for $\mathcal{P}$ given $\mathcal{F}_{i-1}$ if $E_i$ is $\mathcal{F}_i$-measurable and
$$
\mathbb{E}_P[E_i \mid \mathcal{F}_{i-1}] \leq 1
\quad \text{for all } P \in \mathcal{P}.
$$
:::

A conditional e-variable is therefore a valid one-step e-variable once the past has been fixed. Its usefulness is practical as well as conceptual. It allows the analyst to adapt the next step to the data already seen, while preserving validity under the null. In the constructions used here, the rule defining $E_i$ may be chosen using $\mathcal{F}_{i-1}$ and is then evaluated when the new information at step $i$ arrives. This local conditional validity is exactly what makes sequential updating possible [@grunwald2024safe; @ramdas2024hypothesis].

The key payoff from the definition is that conditional e-variables multiply.

::: {#prp-product-construction}
## Product construction [adapted from @grunwald2024safe; @ramdas2024hypothesis]

If $E_1, E_2, \ldots$ are nonnegative random variables such that each $E_i$ is a conditional e-variable for $\mathcal{P}$ given $\mathcal{F}_{i-1}$, then the running product
$$
M_n = \prod_{i=1}^n E_i
$$
is a test supermartingale for $\mathcal{P}$, with $M_0 = 1$, and hence an e-process for $\mathcal{P}$.
:::

::: {.proof}
The empty product gives $M_0 = 1$. For $n \geq 0$, each factor $E_i$ with $i \leq n$ is $\mathcal{F}_i$-measurable and hence $\mathcal{F}_n$-measurable. Therefore, $M_n$ is nonnegative and $\mathcal{F}_n$-measurable, and
$$
\mathbb{E}_P[M_{n+1} \mid \mathcal{F}_n]
=
\mathbb{E}_P[M_n E_{n+1} \mid \mathcal{F}_n]
=
M_n \,\mathbb{E}_P[E_{n+1} \mid \mathcal{F}_n]
\leq M_n.
$$
Thus $(M_n)$ is a test supermartingale, and an e-process by @prp-supermg-eprocess.
:::

This result is useful because it turns a **local** condition, validity one step ahead given the past, into a **global** sequential evidence process. In practice, this is how most concrete e-processes are built. One first constructs a valid one-step factor, then multiplies these factors over time. In that sense, conditional e-variables are not an optional extra piece of theory: they are one of the main construction devices behind sequential safe tests [@grunwald2024safe; @ramdas2024hypothesis].

This construction captures the common sequential backbone behind the tests in @sec-safe-tests. In the **safe z-test**, each factor is a one-step Gaussian likelihood ratio. In the **safe logrank test**, each event contributes a conditional factor based on the current risk set. In the **MLE plug-in** strategy from @sec-composite, where under the null the $X_i$ are iid with density $p$, the estimate $\hat{Q}_{i-1}$ with density $\hat{q}_{i-1}$ is constructed from $\mathcal{F}_{i-1}$, so $X_i$ still has conditional density $p$ given the past and the factor $\hat{q}_{i-1}(X_i)/p(X_i)$ has conditional expectation one. In more elaborate examples such as the **safe t-test** and **safe tests for $2 \times 2$ tables**, the same principle applies, even though the one-step factors may be expressed after coarsening the data or integrating out nuisance parameters [@grunwald2024safe; @turner2024generic; @terschure2024safe; @ramdas2023gametheoretic]. For dependent observations, the denominator would instead need to be the appropriate null conditional density given the past.

The product construction also formalises **optional continuation**, that is, the valid accumulation of evidence across separate stages or studies. Suppose study 1 produces an e-variable $E^{(1)}$ for a null hypothesis $\mathcal{P}$, and let $\mathcal{G}$ denote the information generated by that study, so that $E^{(1)}$ is $\mathcal{G}$-measurable. A second study then collects fresh data and produces $E^{(2)}$, a conditional e-variable for the same null given $\mathcal{G}$. By @prp-product-construction, the product $E^{(1)}E^{(2)}$ is again a valid e-variable for $\mathcal{P}$. Repeating the argument shows that evidence can be accumulated over any number of stages, provided each new factor is conditionally valid given everything that came before, under the same null. In practice this is achieved by basing each new factor on fresh data. This is one of the main advantages of the e-value framework in the optional continuation setting [@grunwald2024safe; @ramdas2024hypothesis].

This is the sequential formalisation of the optional continuation idea already used informally in @sec-safe-tests: one may continue the experiment, or even start a new one, without invalidating the evidence.


## Ville's inequality and anytime-valid testing {#sec-ville}

We now state the theorem that turns the supermartingale property into an anytime-valid testing guarantee. Ville's inequality is the sequential analogue of Markov's inequality: Markov controls threshold crossings at one fixed analysis time, whereas Ville controls the probability of **ever** crossing a threshold along the monitored path [@ramdas2024hypothesis; @ramdas2023gametheoretic].

::: {#thm-ville}
## Ville's inequality [@ville1939; @ramdas2024hypothesis]

Let $(M_n)_{n \geq 0}$ be a nonnegative supermartingale with respect to $(\mathcal{F}_n)$ under $P$, with $\mathbb{E}_P[M_0] \leq 1$. Then for any $c > 0$,
$$
P\!\left(\sup_{n \geq 0} M_n \geq c\right) \leq \frac{1}{c}.
$$
In particular, taking $c = 1/\alpha$, for any $\alpha \in (0,1]$,
$$
P\!\left(\exists\, n \geq 0 : M_n \geq \frac{1}{\alpha}\right) \leq \alpha.
$$
:::

The theorem is stated here in its classical supermartingale form because that is the form proved below. The same crossing bound holds for every e-process by applying its defining stopped-value guarantee to the first-crossing time [@ramdas2024hypothesis, Fact 7.6].

::: {.proof}
Define the stopping time
$$
\tau = \inf\{n \geq 0 : M_n \geq c\},
$$
with $\tau = \infty$ if the set is empty. For any fixed $N$, the stopping time $\tau \wedge N$ is bounded. Since the stopped process $(M_{n \wedge (\tau \wedge N)})_{n \geq 0}$ is again a nonnegative supermartingale,
$$
\mathbb{E}_P[M_{\tau \wedge N}] \leq \mathbb{E}_P[M_0] \leq 1.
$$
Moreover,
$$
M_{\tau \wedge N} \geq c\,\mathbf{1}_{\{\tau \leq N\}},
$$
because on the event $\{\tau \leq N\}$ we have $M_{\tau \wedge N} = M_\tau \geq c$, while on $\{\tau > N\}$ both sides are nonnegative. Taking expectations yields
$$
c \, P(\tau \leq N) \leq \mathbb{E}_P[M_{\tau \wedge N}] \leq 1.
$$
Hence $P(\tau \leq N) \leq 1/c$ for every $N$. Letting $N \to \infty$ and using continuity of probability from below,
$$
P(\tau < \infty) = \lim_{N \to \infty} P(\tau \leq N) \leq \frac{1}{c}.
$$
Since
$$
\{\tau < \infty\}
=
\{\exists\, n \geq 0 : M_n \geq c\},
$$
the argument proves
$$
P\!\left(\exists\, n \geq 0 : M_n \geq c\right)
\leq
\frac{1}{c}.
$$
To obtain the stated supremum bound, fix any $c' \in (0,c)$. If $\sup_{n \geq 0} M_n \geq c$, then $M_n \geq c'$ for some finite $n$, and hence, by the same argument applied at level $c'$,
$$
P\!\left(\sup_{n \geq 0} M_n \geq c\right)
\leq
P\!\left(\exists\, n \geq 0 : M_n \geq c'\right)
\leq
\frac{1}{c'}.
$$
Letting $c' \uparrow c$ gives
$$
P\!\left(\sup_{n \geq 0} M_n \geq c\right)
\leq
\frac{1}{c},
$$
as required.
:::

Applying the theorem with $c = 1/\alpha$ shows that a sequential test which rejects as soon as $M_n \geq 1/\alpha$ has type-I error at most $\alpha$. This is the basic mechanism behind anytime-valid testing with test supermartingales, and hence with the e-processes derived from them [@ramdas2024hypothesis; @grunwald2024safe].

We can now formalise the notion that was used informally throughout @sec-safe-tests.

::: {#def-anytime-valid}
## Anytime-valid test

A sequential test is **anytime-valid** at level $\alpha$ for $\mathcal{P}$ if, for every $P \in \mathcal{P}$,
$$
P\!\left(\exists\, n \geq 1 : \text{the null hypothesis is rejected at time } n\right) \leq \alpha.
$$
:::

This is stronger than fixed-horizon validity: a classical fixed-sample test controls error only at the single sample size for which it was designed, whereas an anytime-valid test controls the probability of **ever** rejecting along the monitored path.

By Ville's inequality, any rule of the form "reject the null whenever $M_n \geq 1/\alpha$" is anytime-valid whenever $(M_n)$ is a test supermartingale. The same guarantee also holds if one evaluates the process at any stopping time $\tau$ with respect to $(\mathcal{F}_n)$ and rejects when $M_\tau \geq 1/\alpha$.

This clarifies the relationship to the group-sequential methods discussed in @sec-safe-logrank. The original O'Brien--Fleming procedure [@obrien1979multiple] fixes the maximum number and spacing of interim analyses. Lan--DeMets alpha-spending [@lan1983discrete] is more flexible: the spending function is specified in advance, but future decision times and the total number of looks need not be known. Anytime-valid tests go further by controlling type-I error uniformly over arbitrary valid stopping times, including stopping rules based on the accumulated evidence itself. The price of that extra flexibility is often reduced fixed-sample efficiency, as seen earlier in @sec-power-cost and @sec-safe-t-power. We already saw the error-control side of this contrast empirically in @fig-safe-z-fpr and @fig-safe-t-fpr.

## Confidence sequences {#sec-confidence-sequences}

The tests in @sec-safe-tests and the theory above answer a binary question: should the null hypothesis be rejected? Estimation asks a richer question: which parameter values remain plausible given the data? Just as e-processes extend e-values to the sequential setting, confidence sequences extend fixed-sample confidence intervals to procedures that remain valid under continuous monitoring.

### Why fixed-sample intervals fail under monitoring {#sec-ci-fail}

A classical $(1-\alpha)$ confidence interval $C_n$ satisfies $P_\theta(\theta \in C_n) \geq 1 - \alpha$ for each fixed, pre-specified sample size $n$, where $P_\theta$ denotes the law of the data when the parameter of interest takes value $\theta$. If the analyst reports $C_n$ only once at that predetermined sample size, this guarantee is valid. But if such intervals are recomputed and inspected repeatedly as data accumulate, this pointwise guarantee does not imply simultaneous coverage over time. In particular, the probability that the interval excludes the true parameter at least once along the monitored path can substantially exceed $\alpha$. As a consequence, fixed-sample intervals also do not in general retain their nominal coverage at arbitrary data-dependent stopping times [@howard2021timeuniform; @ramdas2023gametheoretic, Section 2.8]. This is the estimation analogue of repeated testing: the relevant quantity is no longer coverage at one fixed look, but the probability of ever excluding the truth during monitoring.

To illustrate this distinction, we simulate iid $\mathrm{N}(0, 1)$ observations and compare repeatedly checked fixed-sample Gaussian $z$-intervals with a Gaussian mixture confidence sequence for the mean. For the confidence sequence, we use the half-width formula derived later in @sec-gaussian-cs. The parameter $\rho$ is a tuning parameter of the Gaussian mixture construction; here we set $\rho = 1$ for illustration.

```{r}
#| label: fig-cs-coverage-fail
#| cache: true
#| fig-cap: "Probability of ever excluding the true mean by time $n$ under continuous monitoring, for repeatedly checked fixed-sample Gaussian $z$-intervals and a Gaussian mixture confidence sequence, both nominally at 95\\%. Both quantities are cumulative and therefore nondecreasing. The repeated fixed-sample intervals increasingly fail to cover the true parameter somewhere along the path, whereas the confidence-sequence curve remains below the nominal 5\\% error bound in this simulation. At $n = 1$, the fixed-sample interval has theoretical miss probability exactly $\\alpha$."
#| fig-width: 7
#| fig-height: 4.5

n_max_cov <- 5000
n_sim_cov <- 2000
alpha <- 0.05
rho <- 1

# Precompute half-widths (independent of simulated data)
ci_hw <- qnorm(1 - alpha / 2) / sqrt(1:n_max_cov)
cs_hw <- sqrt(2 * (1 + (1:n_max_cov) * rho) /
              ((1:n_max_cov)^2 * rho) *
              (log(1 / alpha) + 0.5 * log(1 + (1:n_max_cov) * rho)))

set.seed(123)
ever_miss_ci <- numeric(n_max_cov)
ever_miss_cs <- numeric(n_max_cov)

for (sim in 1:n_sim_cov) {
  x <- rnorm(n_max_cov, mean = 0, sd = 1)
  xbar <- cumsum(x) / (1:n_max_cov)

  # At n = 1 the fixed-sample interval
  # misses with probability exactly alpha, which is the reference point
  # against which the later looks degrade
  miss_ci <- which(abs(xbar) > ci_hw)[1]
  miss_cs <- which(abs(xbar) >= cs_hw)[1]

  if (!is.na(miss_ci)) {
    ever_miss_ci[miss_ci:n_max_cov] <-
      ever_miss_ci[miss_ci:n_max_cov] + 1
  }
  if (!is.na(miss_cs)) {
    ever_miss_cs[miss_cs:n_max_cov] <-
      ever_miss_cs[miss_cs:n_max_cov] + 1
  }
}

ever_miss_ci <- ever_miss_ci / n_sim_cov * 100
ever_miss_cs <- ever_miss_cs / n_sim_cov * 100

plot_idx <- unique(c(seq(1, 500, by = 1), seq(500, n_max_cov, by = 10)))

df_cov <- data.frame(
  n = rep(plot_idx, 2),
  pct_miss = c(ever_miss_ci[plot_idx], ever_miss_cs[plot_idx]),
  method = rep(c("Repeated fixed-sample CI",
                 "Confidence sequence"), each = length(plot_idx))
)

ggplot(df_cov, aes(x = n, y = pct_miss, color = method)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  annotate("text", x = n_max_cov * 0.8, y = 8,
           label = "alpha == 5*'%'", parse = TRUE,
           size = 3, color = "grey40") +
  scale_color_manual(values = c(
    "Confidence sequence" = "steelblue",
    "Repeated fixed-sample CI" = "firebrick")) +
  labs(x = "Number of looks / sample size",
       y = "% ever excluding true mean by time n",
       color = NULL) +
  coord_cartesian(ylim = c(0, 65)) +
  theme(legend.position = "bottom")
```

@fig-cs-coverage-fail shows the probability of having excluded the true mean at least once by time $n$. This is the relevant pathwise quantity under continuous monitoring. Repeated fixed-sample intervals, like repeated pointwise tests, do not control it. Confidence sequences are designed precisely to do so: their guarantee is time-uniform, meaning that the true parameter remains inside the sequence of intervals simultaneously over time with probability at least $1-\alpha$.

### Definition and inversion {#sec-cs-definition}

::: {#def-cs}
## Confidence sequence [@lai1976confidence; @howard2021timeuniform; @ramdas2023gametheoretic, Section 2.8]

In a model indexed by $\theta \in \Theta$, a **confidence sequence** at level $1-\alpha$ is a sequence of confidence sets $(C_n)_{n \geq 1}$ such that, for every $\theta \in \Theta$,
$$
P_\theta\!\left(\forall\, n \geq 1 : \theta \in C_n\right) \geq 1 - \alpha.
$$
Equivalently,
$$
P_\theta\!\left(\exists\, n \geq 1 : \theta \notin C_n\right) \leq \alpha.
$$
:::

The phrase "for all $n$" is the crucial difference from an ordinary confidence interval. The parameter must be covered at every sample size simultaneously, not just at one pre-chosen time. As a result, a confidence sequence remains valid under continuous monitoring and therefore also at arbitrary data-dependent stopping times.

Confidence sequences can be constructed by inverting test supermartingales. For each candidate value $\theta_0$, suppose we have a test supermartingale $(M_n^{(\theta_0)})$ for testing $H_0 : \theta = \theta_0$. Define
$$
C_n = \left\{\theta_0 : M_n^{(\theta_0)} < 1/\alpha \right\}.
$$

Then $(C_n)$ is a confidence sequence at level $1-\alpha$. Indeed, if the true parameter value is $\theta$, then
$$
P_\theta\!\left(\exists\, n \geq 1 : \theta \notin C_n\right)
=
P_\theta\!\left(\exists\, n \geq 1 : M_n^{(\theta)} \geq 1/\alpha\right)
\leq \alpha,
$$
where the equality restates the definition of $C_n$, and the inequality is the second display of @thm-ville applied to the test supermartingale associated with the true parameter value. Thus, the true parameter can leave the confidence set at some time only if its own test supermartingale ever crosses the threshold $1/\alpha$, and the probability of this happening is at most $\alpha$.

Note that $(C_n)$ need not be nested: because the e-process value $M_n^{(\theta_0)}$ can rise above $1/\alpha$ at one time and fall below it later, a candidate $\theta_0$ can leave $C_n$ and subsequently re-enter $C_{n'}$ for some $n' > n$. For applications where a monotonically shrinking sequence is desired, one can take the running intersection
$$
\widetilde{C}_n
=
\bigcap_{i \leq n} C_i
=
\left\{\theta_0 : \max_{i \leq n} M_i^{(\theta_0)} < 1/\alpha \right\},
$$
which is nested by construction and inherits the same simultaneous coverage guarantee. In the Gaussian example below, the pointwise sets $C_n$ are the more convenient objects to derive and plot.

### A Gaussian mixture confidence sequence {#sec-gaussian-cs}

To make the construction concrete, we derive a confidence sequence for the mean of Gaussian observations with known variance. This uses the same method of mixtures as in @sec-composite, here with a Gaussian mixing distribution, now applied pointwise to each candidate value $\theta_0$.

Suppose $X_1, X_2, \ldots \overset{\text{iid}}{\sim} \mathrm{N}(\theta, 1)$ with unknown mean $\theta$. For a candidate value $\theta_0$, define
$$
S_n^{(\theta_0)} = \sum_{i=1}^n (X_i - \theta_0) = n(\bar{X}_n - \theta_0).
$$
For a fixed local alternative $\theta_0 + \lambda$, let
$$
L_n^{(\theta_0,\lambda)} = \exp\!\left(\lambda S_n^{(\theta_0)} - \frac{n\lambda^2}{2}\right).
$$
This is the likelihood ratio of $\mathrm{N}(\theta_0+\lambda,1)$ against $\mathrm{N}(\theta_0,1)$ after $n$ observations. Setting $S_0^{(\theta_0)} = 0$, as the empty sum requires, gives $L_0^{(\theta_0,\lambda)} = 1$, so the process $(L_n^{(\theta_0,\lambda)})_{n \geq 0}$ is a test martingale under $P_{\theta_0}$ for each fixed $\lambda$.

The statistic $S_n^{(\theta_0)}$ measures the cumulative deviation of the data from the candidate value $\theta_0$. If $\bar{X}_n$ stays close to $\theta_0$, then $S_n^{(\theta_0)}$ remains small and little evidence accumulates against that candidate. If $\bar{X}_n$ drifts away from $\theta_0$, then the evidence against $\theta_0$ increases.

Since the direction and size of the deviation from $\theta_0$ are unknown, we average this likelihood ratio over a symmetric Gaussian mixing distribution $\lambda \sim \mathrm{N}(0, \rho)$ with variance $\rho > 0$, where $\rho$ is a fixed tuning parameter chosen in advance. Because the mixing distribution is symmetric around zero, positive and negative deviations from $\theta_0$ are handled within the same construction, yielding a two-sided mixture [see @ramdas2023gametheoretic, Section 5.1; @howard2021timeuniform]:
$$
M_n^{(\theta_0)} = \int_{-\infty}^{\infty} L_n^{(\theta_0,\lambda)} \frac{1}{\sqrt{2\pi\rho}}\, \exp\!\left(-\frac{\lambda^2}{2\rho}\right) d\lambda.
$$

This is the same mixture strategy used for composite hypotheses in @sec-composite, now applied pointwise for each candidate value $\theta_0$. The tuning parameter $\rho$ determines how widely the mixture spreads its mass over possible deviations from $\theta_0$. In Section 5.1 of @ramdas2023gametheoretic the centred Gaussian mixing distribution is written as $\mathrm{N}(0, \tilde{\rho}^{\,2})$, so $\tilde{\rho}^{\,2} = \rho$ in the notation used here.

Combining the quadratic terms in $\lambda$ gives
$$
M_n^{(\theta_0)} = \int_{-\infty}^{\infty} \frac{1}{\sqrt{2\pi\rho}}\, \exp\!\left(\lambda S_n^{(\theta_0)} - \frac{1}{2}\!\left(n + \frac{1}{\rho}\right)\lambda^2\right) d\lambda.
$$
Write $\sigma_\ast^2 = 1/(n + 1/\rho)$ and $\mu^\ast = \sigma_\ast^2\, S_n^{(\theta_0)}$. Then the exponent in the integrand can be rewritten as
$$
\lambda S_n^{(\theta_0)} - \frac{1}{2\sigma_\ast^2}\,\lambda^2
\;=\;
-\frac{1}{2\sigma_\ast^2}\bigl(\lambda - \mu^\ast\bigr)^2
\;+\; \frac{\bigl(S_n^{(\theta_0)}\bigr)^2}{2(n + 1/\rho)},
$$
so the integral becomes
$$
M_n^{(\theta_0)}
\;=\;
\frac{1}{\sqrt{2\pi\rho}}\,
\exp\!\left(\frac{\bigl(S_n^{(\theta_0)}\bigr)^2}{2(n + 1/\rho)}\right)
\int_{-\infty}^{\infty}
\exp\!\left(-\frac{(\lambda - \mu^\ast)^2}{2\sigma_\ast^2}\right) d\lambda.
$$
The remaining integral is that of an unnormalised Gaussian with variance $\sigma_\ast^2 = 1/(n+1/\rho)$, so it evaluates to $\sqrt{2\pi\sigma_\ast^2} = \sqrt{2\pi/(n+1/\rho)}$. Multiplying by the prefactor $1/\sqrt{2\pi\rho}$ gives
$$
\frac{1}{\sqrt{2\pi\rho}} \cdot \sqrt{\frac{2\pi}{n + 1/\rho}}
\;=\; \frac{1}{\sqrt{\rho\,(n + 1/\rho)}}
\;=\; \frac{1}{\sqrt{1 + n\rho}}.
$$
Thus
$$
M_n^{(\theta_0)} = \frac{1}{\sqrt{1 + n\rho}}\, \exp\!\left(\frac{\bigl(S_n^{(\theta_0)}\bigr)^2}{2(n + 1/\rho)}\right).
$$
Since $1/(n+1/\rho) = \rho/(1+n\rho)$, this is equivalently
$$
M_n^{(\theta_0)} = \frac{1}{\sqrt{1 + n\rho}}\, \exp\!\left(\frac{\rho\, \bigl(S_n^{(\theta_0)}\bigr)^2}{2(1 + n\rho)}\right).
$$

Substituting $S_n^{(\theta_0)} = n(\bar{X}_n - \theta_0)$ gives the closed form
$$
M_n^{(\theta_0)} = \frac{1}{\sqrt{1 + n\rho}}\, \exp\!\left(\frac{n^2 \rho}{2(1 + n\rho)}\, (\bar{X}_n - \theta_0)^2\right).
$$

This is in fact a test martingale under $\theta = \theta_0$. For each fixed $\lambda$, the process $(L_n^{(\theta_0,\lambda)})_{n \geq 0}$ is a test martingale under $P_{\theta_0}$, and integrating these processes against a fixed proper mixing distribution preserves the martingale property, by Tonelli's theorem for conditional expectations, which applies because the integrand is nonnegative. It is therefore also an e-process under $P_{\theta_0}$ by @prp-supermg-eprocess. Since $M_n^{(\theta_0)}$ increases with $|\bar{X}_n - \theta_0|$, candidate means further from the running sample mean yield larger e-values. Thus, when we invert the process, the surviving values of $\theta_0$ will be precisely those that remain sufficiently close to $\bar{X}_n$.

To invert the e-process at time $n$, require $M_n^{(\theta_0)} < 1/\alpha$ and take logarithms:
$$
-\frac{1}{2}\log(1 + n\rho) + \frac{n^2\rho}{2(1 + n\rho)}\,(\bar{X}_n - \theta_0)^2 < \log(1/\alpha).
$$
Rearranging gives
$$
(\bar{X}_n - \theta_0)^2 < \frac{2(1 + n\rho)}{n^2 \rho}\left[\log(1/\alpha) + \frac{1}{2}\log(1 + n\rho)\right].
$$
Thus the half-width is
$$
w_n = \sqrt{\frac{2(1 + n\rho)}{n^2 \rho}\left[\log\!\left(\frac{1}{\alpha}\right) + \frac{1}{2}\log(1 + n\rho)\right]}
$$ {#eq-cs-halfwidth}


Equivalently,
$$
C_n = (\bar{X}_n - w_n,\; \bar{X}_n + w_n).
$$
No $\alpha/2$ appears here because the symmetric Gaussian mixture already defines a two-sided test: positive and negative deviations from $\theta_0$ are handled within the same e-process, so inverting the threshold $1/\alpha$ directly yields a two-sided confidence sequence.

The form of this interval is familiar, but its width differs from that of a classical fixed-sample Gaussian interval. A standard fixed-sample interval has half-width proportional to $1/\sqrt{n}$, whereas the additional $\log(1+n\rho)$ term here slows that decay. The consequences are worked out in @sec-cs-width.

This half-width formula will be used directly in the R implementation below.

### Confidence sequences in R {#sec-cs-r}

We now implement the Gaussian mixture confidence sequence and compare it with the ordinary fixed-sample Gaussian confidence interval with known variance, evaluated at each sample size $n$.

```{r}
#| label: fig-cs-path
#| fig-cap: "A Gaussian mixture confidence sequence (shaded blue band) and fixed-sample 95\\% Gaussian confidence intervals evaluated at each sample size (red dashed lines) for the same data stream. The grey solid line is the running sample mean $\\bar{X}_n$ and the true mean is $\\theta = 0.5$ (grey dotted line). The tuning parameter is fixed at $\\rho = 1$ for illustration. Display starts at $n = 5$ for readability."
#| fig-width: 8
#| fig-height: 5
theta_true <- 0.5
n_max <- 400
alpha <- 0.05
rho <- 1

set.seed(42)
x <- rnorm(n_max, mean = theta_true, sd = 1)
xbar <- cumsum(x) / (1:n_max)

cs_hw <- sqrt(2 * (1 + (1:n_max) * rho) / ((1:n_max)^2 * rho) *
              (log(1 / alpha) + 0.5 * log(1 + (1:n_max) * rho)))

ci_hw <- qnorm(1 - alpha / 2) / sqrt(1:n_max)

# Start plotting at n = 5 to avoid visually extreme early-sample behaviour
path_idx <- 5:n_max
df_cs <- data.frame(
  n = path_idx,
  xbar = xbar[path_idx],
  cs_lo = xbar[path_idx] - cs_hw[path_idx],
  cs_hi = xbar[path_idx] + cs_hw[path_idx],
  ci_lo = xbar[path_idx] - ci_hw[path_idx],
  ci_hi = xbar[path_idx] + ci_hw[path_idx]
)

ggplot(df_cs, aes(x = n)) +
  geom_ribbon(aes(ymin = cs_lo, ymax = cs_hi),
              fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = cs_lo), color = "steelblue", linewidth = 0.5) +
  geom_line(aes(y = cs_hi), color = "steelblue", linewidth = 0.5) +
  geom_line(aes(y = ci_lo), color = "firebrick",
            linetype = "dashed", linewidth = 0.4) +
  geom_line(aes(y = ci_hi), color = "firebrick",
            linetype = "dashed", linewidth = 0.4) +
  geom_line(aes(y = xbar), color = "grey30", linewidth = 0.4) +
  geom_hline(yintercept = theta_true, linetype = "dotted",
             color = "grey50") +
  annotate("text", x = n_max - 5, y = theta_true + 0.12,
           label = "theta == 0.5", parse = TRUE,
           hjust = 1, size = 3, color = "grey50") +
  labs(x = "Sample size",
       y = "Mean and interval bounds")
```

@fig-cs-path shows one realised Gaussian data stream. At early sample sizes, the confidence sequence is noticeably wider than the fixed-sample interval evaluated at the same $n$: this is the cost of simultaneous validity over time. As the sample size grows, both intervals contract and the absolute gap between them narrows, but the confidence sequence remains wider throughout, and the ratio of the two half-widths eventually grows slowly with $n$, as @sec-cs-width makes precise.

The blue band retains its coverage guarantee under continuous monitoring, whereas the red dashed intervals have 95% coverage only pointwise at each fixed sample size.

For illustration, the figure fixes the tuning parameter at $\rho = 1$. The simultaneous coverage guarantee holds for any fixed choice of $\rho > 0$; changing $\rho$ affects the width profile of the sequence over time, but not its validity. In that sense, $\rho$ is a tuning parameter, analogous to the design choices used for safe tests in @sec-safe-tests.


### Width of the Gaussian mixture confidence sequence {#sec-cs-width}

The Gaussian mixture confidence sequence is wider than the corresponding fixed-sample Gaussian confidence interval because it must remain valid under continuous monitoring, not just at one pre-specified sample size. This extra width is the price of time-uniform validity. Starting from the exact half-width in @eq-cs-halfwidth, a large-$n$ approximation uses $1 + n\rho \approx n\rho$, which gives
$$
w_n \;\approx\; \sqrt{\frac{2\bigl[\log(1/\alpha) + \tfrac{1}{2}\log(n\rho)\bigr]}{n}}
\;\asymp\;
\sqrt{\frac{\log n}{n}}.
$$
Here $f(n) \asymp g(n)$ means that the ratio $f(n)/g(n)$ is bounded above and below by positive constants for large $n$. Since $\log(1/\alpha)$ and $\tfrac{1}{2}\log\rho$ are constant in $n$, the dominant term for large $n$ is $\tfrac{1}{2}\log n$, so the half-width shrinks to zero, but more slowly than the fixed-sample $n^{-1/2}$ rate.

By contrast, the fixed-sample Gaussian confidence interval has half-width
$$
z_{1-\alpha/2}/\sqrt{n} \;\asymp\; 1/\sqrt{n}.
$$
Thus, this Gaussian mixture confidence sequence incurs an additional factor of order $\sqrt{\log n}$ relative to the fixed-sample interval. The parameter $\rho$ is a tuning parameter of the Gaussian mixture construction: it affects the finite-sample width, but not the asymptotic rate $\sqrt{\log n / n}$. Although the extra $\sqrt{\log n}$ factor diverges, it does so very slowly, so the resulting widening is limited over a broad range of sample sizes.

Sharper confidence sequences with smaller time-uniform penalties are available [see, e.g., @howard2021timeuniform], but the Gaussian mixture construction is especially useful here because it is explicit, easy to derive from an e-process, and easy to invert. In a sequential setting such as a clinical trial with repeated interim looks, this means that the price of being allowed to monitor the interval continuously is a modest widening relative to the interval one would report at a single fixed analysis.

```{r}
#| label: fig-cs-width-ratio
#| fig-cap: "Ratio of the Gaussian mixture confidence-sequence half-width to the fixed-sample Gaussian confidence-interval half-width for three values of $\\rho$. The widening factor grows slowly with $n$, illustrating the cost of requiring validity at all stopping times."
#| fig-width: 6
#| fig-height: 4
nn <- 20:2000
rho_values <- c(0.5, 1, 2)

df_ratio <- do.call(rbind, lapply(rho_values, function(rho) {
  cs_w <- sqrt(2 * (1 + nn * rho) / (nn^2 * rho) *
               (log(1 / alpha) + 0.5 * log(1 + nn * rho)))
  ci_w <- qnorm(1 - alpha / 2) / sqrt(nn)
  data.frame(
    n = nn,
    ratio = cs_w / ci_w,
    rho = paste0("rho == ", rho)
  )
}))

ggplot(df_ratio, aes(x = n, y = ratio, color = rho)) +
  geom_line(linewidth = 0.7) +
  scale_color_discrete(labels = function(l) parse(text = l)) +
  expand_limits(y = 1) +
  labs(x = "Sample size",
       y = "CS half-width / CI half-width",
       color = NULL) +
  theme(legend.position = "bottom")
```

@fig-cs-width-ratio shows that the confidence-sequence half-width exceeds the fixed-sample half-width by a factor that increases only slowly with $n$. Over the plotted range, the ratio never exceeds `r round(max(df_ratio$ratio), 2)` for any of the three values of $\rho$, illustrating that the cost of time-uniform validity is visible but not dramatic.


### Beyond the Gaussian case {#sec-beyond-gaussian}

The Gaussian mixture confidence sequence above relied on the same simplifying assumption as the safe z-test in @sec-safe-z: the variance was treated as known. In more realistic settings, the same broad strategy still applies, but the construction of the underlying e-process becomes more delicate.

For a normal mean with unknown variance, the nuisance scale parameter prevents the simple known-variance likelihood-ratio construction from carrying over unchanged. The right-Haar-prior ideas that already appeared in the safe t-test in @sec-safe-t are also relevant here, and Lai's classical confidence-sequence construction can be understood from this perspective [@lai1976confidence]. However, as @wang2025anytimevalid explain, Lai's construction combines an improper right-Haar mixture over the scale parameter with an improper flat mixture over the mean $\mu$. The resulting process is an extended nonnegative supermartingale rather than an ordinary integrable martingale, and therefore is not an e-process. They present two valid alternatives: a test martingale in a reduced scale-invariant filtration and an e-process in the canonical data filtration. We refer the reader to their paper for the full constructions.

Beyond Gaussian models, sub-Gaussian methods apply to observations that need not be bounded, while empirical-Bernstein methods provide variance-adaptive nonparametric confidence sequences for bounded observations [@howard2021timeuniform; @waudbysmith2024estimating].

For the e-process-based constructions described above, the unifying principle is the same: construct a family of e-processes indexed by the candidate parameter value, obtain an anytime-valid test for each candidate by thresholding at $1/\alpha$, and then invert that family of tests to obtain a confidence sequence.

## The common backbone of safe tests {#sec-common-backbone}

We can now reinterpret the safe tests from @sec-safe-tests as instances of a common sequential template:

1. **Construct conditional e-variables.** At each time step, form a one-step e-variable that is conditionally valid given the past.
2. **Multiply them.** By @prp-product-construction, the running product is a test supermartingale.
3. **Apply Ville's inequality.** Thresholding this test supermartingale at $1/\alpha$ gives an anytime-valid test.
4. **Invert.** If the resulting family of test supermartingales is indexed by a candidate parameter value, inverting the corresponding anytime-valid tests yields a confidence sequence.

This template is cleanest when one-step conditional e-variables are available directly in the chosen filtration. Some important constructions are slightly less direct: they require a suitable coarsening of the filtration or a different representation of the data before the sequential validity argument takes this product form.

Across the safe tests, the common backbone is therefore the same, while the main modelling work lies in step 1. In the safe z-test, the one-step factor is a Gaussian likelihood ratio; see @sec-safe-z. In the safe t-test, the construction only becomes valid after handling the nuisance scale parameter through the right-Haar-prior approach and its associated invariant structure [@grunwald2024safe]; see @sec-safe-t. In the safe $2 \times 2$ test, the factor is based on the beta-binomial construction of @turner2024generic; see @sec-safe-2x2. In the safe logrank test, it is built from conditional risk-set contributions [@terschure2024safe]; see @sec-safe-logrank. Once an appropriate conditionally valid factor has been identified in the relevant filtration, the remaining supermartingale argument and the use of Ville's inequality are generic.

This is also, in broad terms, the practical architecture of the `safestats` package [@turner2022safestats]. The `designSafe*` functions specify the design parameters of the safe procedure, while the `safe*Test` functions update the running evidence and compare it with the threshold $1/\alpha$.

`Figure~\ref{fig-common-backbone}`{=latex} summarises this four-step template.

```{=latex}
\begin{figure}[H]
\centering
\begin{tikzpicture}[
  >={Stealth[length=2.2mm]},
  node distance=8mm and 14mm,
  every node/.style={font=\small}
]

\node[backbone-step] (cond) {%
  \textbf{1.\ Conditional e-variables}\\
  one-step factor $E_i$ with $\mathbb{E}_P[E_i \mid \mathcal{F}_{i-1}] \leq 1$
};
\node[backbone-step, below=of cond] (prod) {%
  \textbf{2.\ Product construction}\\
  $M_n = \prod_{i=1}^{n} E_i$ is a test supermartingale
};
\node[backbone-step, below=of prod] (ville) {%
  \textbf{3.\ Ville's inequality}\\
  reject when $M_n \geq 1/\alpha$: anytime-valid test
};
\node[backbone-step, below=of ville] (inv) {%
  \textbf{4.\ Inversion}\\
  parameter-indexed family $(M_n^{(\theta_0)})$ $\Rightarrow$ confidence sequence
};

\node[backbone-tests, right=22mm of cond] (instances) {%
  \textbf{Problem-specific:}\\[2pt]
  $\bullet$\ safe $z$: Gaussian LR\\
  $\bullet$\ safe $t$: right-Haar / invariant\\
  \hspace*{1em}construction\\
  $\bullet$\ safe $2{\times}2$: beta-binomial\\
  $\bullet$\ safe logrank: risk-set factor
};

\draw[backbone-arr] (cond) -- (prod);
\draw[backbone-arr] (prod) -- (ville);
\draw[backbone-arr] (ville) -- (inv);
\draw[backbone-sidearr] (instances) -- (cond);

\end{tikzpicture}
\caption{Common sequential backbone behind the safe tests of Chapter~3. Steps 2--4 are generic; only the construction of the one-step factor in step~1 is problem-specific. Inversion to a confidence sequence applies when a parameter-indexed family of such processes is available.}
\label{fig-common-backbone}
\end{figure}
```

## Summary {#sec-ch4-summary}

This chapter developed the sequential machinery that underlies the safe tests in @sec-safe-tests.

1. **Test supermartingales and e-processes** extend e-values from single analyses to evolving evidence streams. In this chapter, test supermartingales were the main construction; by @prp-supermg-eprocess, every test supermartingale is also an e-process, meaning its value at every stopping time is a valid e-value.

2. **Conditional e-variables and the product construction** explain how sequential evidence is built. At each step, one chooses a conditionally valid e-variable based on the past, and the running product remains valid under the null. This is the basic mechanism behind adaptive sequential monitoring and optional continuation.

3. **Ville's inequality** is the key theorem behind anytime-valid testing. For a test supermartingale, it guarantees that the probability of ever crossing the threshold $1/\alpha$ is at most $\alpha$, with no restriction on when the analyst looks. Separately, the value of the process at a stopping time with respect to the chosen filtration is itself a valid e-value.

4. **Confidence sequences** are the estimation analogue of anytime-valid tests. They are obtained by inverting a family of anytime-valid tests derived from e-processes and provide simultaneous coverage over all sample sizes.

5. **Safe tests share a common sequential backbone.** What changes from one testing problem to another is the construction of an appropriate e-process or test supermartingale, often via conditional e-variables and sometimes after a suitable coarsening or re-expression of the data; once such a process is available, Ville-based validity is generic.

In the next chapter, we turn from a single hypothesis to many. The same e-value framework extends naturally to multiple testing, where methods such as eBH and the e-Partitioning Principle provide tools for false discovery rate control with e-values.